\documentclass[10pt]{beamer}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm,latexsym}
\usepackage{mathrsfs}
\usepackage[all]{xypic}
\usepackage{tikz}
\usetikzlibrary{arrows,shapes,automata}

\newcommand{\cat}[1]{\mathscr{#1}}
\newcommand{\lcat}[1]{\mathbf{#1}}
\newcommand{\C}{\cat{C}}
\newcommand{\D}{\cat{D}}
\renewcommand{\L}{\cat{L}}
\newcommand{\comp}{\circ}
\newcommand{\size}[1]{\lVert#1\rVert}
\newcommand{\sizein}[1]{\size{#1}^\mathrm{in}}
\newcommand{\sizeout}[1]{\size{#1}_\mathrm{out}}
\DeclareMathOperator{\ob}{ob}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Id}{Id}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Complex}{\mathbb{C}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\K}{\mathbb{K}}
\newcommand{\LK}{\mathbb{L}}
\newcommand{\ra}{\rightarrow}
\newcommand{\la}{\leftarrow}
\DeclareMathOperator{\Time}{t}
\DeclareMathOperator{\Space}{s}
\DeclareMathOperator{\coTime}{co-t}
\DeclareMathOperator{\coSpace}{co-s}
\DeclareMathOperator{\Par}{Par}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\rev}{rev}
\newcommand{\computes}{\vdash}


\mode<presentation>{%
  \usetheme[]{Madrid}
  \usecolortheme{seahorse}
  \usecolortheme{rose}
  \usefonttheme[onlymath]{serif}
}


\title{On the transposition of black box matrices}
\author[L. De Feo]{L.~De~Feo\\{\footnotesize(joint work with É.~Schost)}}
\institute[TANC, LIX]{Projet TANC, LIX, École Polytechnique}
\date[Typical, February 11, 2010]{Séminaire TypiCal\\February 11, 2010}

%% \AtBeginSection[]
%% {
%%   \begin{frame}<beamer>
%%     \frametitle{Plan}
%%     \tableofcontents[currentsection]
%%   \end{frame}
%% }


\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\section{Black-box model}

\begin{frame}
  \frametitle{The power iteration}

\end{frame}

%%

\begin{frame}
  \frametitle{From numerical algorithms to exact algorithms}

\end{frame}

%%

\begin{frame}
  \frametitle{From sparse to black-box matrices}

\end{frame}

%%
%%

\section{Algebraic complexity}

\begin{frame}
  \frametitle{Measuring the complexity of black-box algorithms}

\end{frame}

%%

\begin{frame}
  \frametitle{Arithmetic circuits}

\end{frame}

%%

\begin{frame}
  \frametitle{Arithmetic Circuits: uniform vs. non-uniform}

\end{frame}

%%
%%

\section{The transposition theorem}

\begin{frame}
  \frametitle{Transposition of an arithmetic circuit}

\end{frame}

%%

\begin{frame}
  \frametitle{History of the transposition theorem}

\end{frame}

%%

\begin{frame}
  \frametitle{From non-uniform to uniform circuits?}

\end{frame}

%%

\begin{frame}
  \frametitle{The case of multiplication}

\end{frame}

%%

\begin{frame}
  \frametitle{The parameter space: \texttt{linear} and \texttt{const}
    variables}

\end{frame}

%%

\begin{frame}
  \frametitle{Formal definition of a \emph{linear program}}

\end{frame}

%%
%%

\section{A transposable language}

\begin{frame}
  \frametitle{TransAL: the Transposable Algebraic Language}

\end{frame}

%%

\begin{frame}
  \frametitle{Transposing a simple code}

\end{frame}

%%

\begin{frame}
  \frametitle{Automatic transposition: the general algorithm}

\end{frame}

%%

\begin{frame}
  \frametitle{Non-linear prediction and tail recursion}

\end{frame}

%%
%%

\section{Constness inference system}

\begin{frame}
  \frametitle{Automatic deduction of constness, a simple idea}

\end{frame}

%%

\begin{frame}
  \frametitle{The problem, relationship with generically quantified types}

\end{frame}

%%

\begin{frame}
  \frametitle{A simple algorithm}

\end{frame}

%%
%%

\section{Transalpyne}

\begin{frame}
  \frametitle{A Python implementation of TransAL}

\end{frame}

%%
%%

\section{A categorical point of view}

\begin{frame}
  \frametitle{Arrows, not circuits !}

\end{frame}

%%

\begin{frame}
  \frametitle{Transposition as a contravariant functor}

\end{frame}

%%

\begin{frame}
  \frametitle{Some trivial applications}

\end{frame}

%%

\begin{frame}
  \frametitle{Some interesting problems}

\end{frame}

%%
%%

\begin{frame}
  \frametitle{Bibliography}

  \begin{thebibliography}{1}
  
  \bibitem[B\"urgisser, Clausen, Shokrollahi]{BCS}P.~Bürgisser, M.~Clausen \& M.~A.~Shokrollahi,
    \newblock \emph{Algebraic Complexity Theory}.
    \newblock Springer, 1997.
    
  \bibitem[Bostan, Lercerf, Schost '03]{BLS03}A.~Bostan, G.~Lecerf \& E.~Schost,
    \newblock Tellegen's Principle into Practice.
    \newblock \emph{Proceedings of ISAAC 2003}.

  \bibitem[D.F., Schost '09]{DFS09}
    L.~De~Feo and {\'E}.~Schost.
    \newblock Fast arithmetics in Artin-Schreier towers over finite fields. 
    \newblock In \emph{ISSAC'09}, pages 127--134. ACM, 2010.

  \bibitem[Kaltofen '00]{Ka2k}
    E.~Kaltofen.
    \newblock Challenges of symbolic computation: my
    favorite open problems.
    \newblock \emph{Journal of Symbolic Computation}, 29(6):891--919, 2000.

  \bibitem[Kaltofen, Trager '90]{KaTr90}
    E.~Kaltofen, B.~M.~Trager.
    \newblock Computing with polynomials given by black boxes for their
    evaluations: greatest common divisors, factorization, separation of 
    numerators and denominators.
    \newblock \emph{J. Symb. Comput.}, 9:301--320, 1990.

  \end{thebibliography}
\end{frame}

\begin{frame}
  \frametitle{Bibliography}

  \begin{thebibliography}{1}
  
  \bibitem[Page, Brin, Motwani, Winograd '99]{google}
    L.~Page, S.~Brin, R.~Motwani and T.~Winograd.
    \newblock The PageRank Citation Ranking: Bringing Order to the Web.
    \newblock \emph{Technical Report}, Stanford InfoLab, 1999,
    \url{http://ilpubs.stanford.edu:8090/422/}.

  \bibitem[Pascal, Schost '06]{PS06}
    C.~Pascal and \'E.~Schost.
    \newblock Change of order for bivariate triangular sets.
    \newblock In \emph{ISSAC'06}, pages 277--284. ACM, 2006.
    
  \bibitem[Shoup '95]{Sho95}
    V.~Shoup.
    \newblock A new polynomial factorization algorithm and its implementation.
    \newblock \emph{J. Symb. Comp.}, 20(4):363--397, 1995.
    
  \bibitem[Wiedemann '86]{Wie86}
    D.~H.~Wiedemann.
    \newblock Solving Sparse Linear Equations Over Finite Fields.
    \newblock \emph{IEEE Trans. Inf. Theory}, vol. IT-32:54--62, 1986.

  \end{thebibliography}
\end{frame}

\end{document}





%%% Local Variables: 
%%% mode:flyspell
%%% ispell-local-dictionary:"british"
%%% mode: TeX-PDF
%%% TeX-master: t
%%% End: 
%
% LocalWords:  Tellegen